### 2015-12-01 doc/thesis/intro.tex
Wrote some more

In the case of a rooted tree with branch lengths in calendar time, the internal
nodes can be mapped onto a timeline by using the tips of the phylogeny, which
usually map to the present day, as a reference point. These \defn{branching
times} correspond to the times at which speciation events occurred. In
phylogenies relating animal species, speciation most often occurs by an
\defn{allopatric} process, where a sub-population of organisms is
geographically isolated from the original population. Over time, the two
populations diverge genetically, eventually resulting in two distinct species.

Certain viruses, particularly RNA viruses such as the human immuno-deficiency
virus (HIV), evolve rapidy enough that populations can accumulate measurable
genetic changes on time scales of months or years. Transmission of the virus
from one host to another closely mirrors the allopatric speciation process.
Upon transmission, a sub-population of virions from the donor host is
geographically isolated in the recpient, followed by genetic divergence of the
two hosts' infections. Therefore, the branching time relating the two hosts'
viral populations in a time-scaled phylogeny corresponds to the time of
transmission. This correspondence can be quite rough, and is subject to some
important limitations which are discussed below. However, even taking these
limitations into account, the relationship between branching times in the
phylogeny and transmission times in the population theoretically implies that
the evolutionary history of the virus is informative about its epidemiology. In
other words, information about transmission patterns can be inferred from the
viral phylogeny, complementing more traditional survey-based epidemiological
methods. The tools and techniques for making such inferences are referred to as
\defn{phylodynamic} methods.

@@

A \defn{contact network} is a graphical representation of a population over
which an epidemic can spread. Contact networks consist of nodes and edges,
where the nodes represent people, and edges represent contacts. An edge between
two nodes indicate that transmission is possible between the two corresponding
individuals. For example, if we are considering an HIV transmission network,
the edges may represent sexual contacts or individuals sharing injection drug
paraphernalia. For an airborne pathogen such as influenza, edges may simply
indicate that the two people were in the same room. The edges in a contact
network may be undirected, indicating that all contacts could possibly transmit
to each other, or directed, indicating that transmission risk is one-way. It is
possible to simulate an undirected network using directed edges by simply
including two directed edges, one in either direction, for each single directed
edge. Therefore, we exclusively consider directed contact networks here.

>>>>

In the case of a rooted tree with branch lengths in calendar time, the internal
nodes can be mapped onto a timeline by using the tips of the phylogeny, which
usually map to the present day, as a reference point. These \defn{branching
times} correspond to the times at which speciation events
occurred~\autocite{nee1992tempo}. In phylogenies relating animal species,
speciation most often occurs by an \defn{allopatric} process, where a
sub-population of organisms is geographically isolated from the original
population [a reference to Coyne \& Orr will go here once I skim the relevant
bits of the book]. Over time, the two populations diverge genetically,
eventually resulting in two distinct species.

Many RNA viruses, such as the human immuno-deficiency virus (HIV), evolve
rapidy enough that populations can accumulate measurable genetic changes on
time scales of months or years~\autocite{drummond2003measurably}. Transmission
of the virus from one host to another closely mirrors the allopatric speciation
process. Upon transmission, a sub-population of virions from the donor host is
geographically isolated in the recpient, followed by genetic divergence of the
two hosts' infections. Therefore, the branching time relating the two hosts'
viral populations in a time-scaled phylogeny corresponds to the time of
transmission. This correspondence can be quite rough, and is subject to some
important limitations which are discussed below. However, even taking these
limitations into account, the relationship between branching times in the
phylogeny and transmission times in the population theoretically implies that
the evolutionary history of the virus is informative about its epidemiology. In
other words, information about transmission patterns can be inferred from the
viral phylogeny, complementing more traditional survey-based epidemiological
methods. The tools and techniques for making such inferences are referred to as
\defn{phylodynamic} methods.

@@

A \defn{contact network} is a graphical representation of a population over
which an epidemic can spread. Contact networks consist of nodes and edges,
where the nodes represent people, and edges represent contacts. An edge between
two nodes indicate that transmission is possible between the two corresponding
individuals. For example, if we are considering an HIV transmission network,
the edges may represent sexual contacts or individuals sharing injection drug
paraphernalia. For an airborne pathogen such as influenza, an edge may simply
indicate that the two people were in the same room. The edges in a contact
network may be undirected, indicating that all contacts could possibly transmit
to each other, or directed, indicating that transmission risk is one-way. It is
possible to simulate an undirected network using directed edges by simply
including two directed edges, one in either direction, for each single directed
edge. Therefore, we exclusively consider directed contact networks here.

Contact networks have been studied extensively in a variety of social,
economic, and public health contexts. However, investigation of the effects of
contact network structures on viral phylogenies have been much more limited.
Existing studies have to focused on a few broad classes of contact network, and
how phylogenies deriving from these networks have
differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
robinson2013dynamics}. To demonstrate our method, we focus on the same three
contact network types as studied in~\autocite{leventhal2012inferring}, namely
Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
graphs. ER graphs, also known as random graphs, are the simplest generative
graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
which gives the probability of any edge occuring in the network. BA
graphs~\autocite{barabasi1999emergence} model a preferential attachment process
where nodes with high degree tend to attract more connections. They have two
parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
nodes with $m$ outgoing edges to the network. An existing node $v$ in the
network is chosen to be the endpoint of these new edges with probability
proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
graphs~\autocite{watts1998collective}, also known as small-world graphs, have
two parameters, $k$ and $p$. They are constructed by creating a ring lattice
with $k$ edges per node, and then randomly rewiring each edge with probability
$p$.

\subsection{Approximate Bayesian computation}

Consider a model $M$, with parameters $\theta$, which we wish to fit to some
observed data $D$. By ``fit'', we often mean that we want to find particular
values $\hat{\theta}$ for the parameters which optimize the likelihood of our
data given those parameters and the model,
\[
  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
\]
This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
being unsed to refer to a probabilty \emph{density} rather than a true
probability. Alternative to maximum likelihood, we may be interested less in a
point estimate and more in the posterior distribution of possible values of
$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
below, but for the moment, for illustrative purposes, we restrict our attention
to the maximum likelihood problem.

If the model we are fitting is sufficiently simple, it may be possible to
calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
analytic maximum likeilhood solutions, but if the likelihood any set of
parameters can be calculated up to a normalizing constant, then
${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
optimization strategies exist, the choice of which to use depending on the
complexity of the model and whether or not we have access to the gradients of
the likelihood function with respect to each of the parameters. The majority of
modelling problems fit into this category, and numerical optimization is
well-developed and extremely widely used.

However, there are some cases, often when the observed data is of a complex
type, that explicitly calculating the likelihood of some observed data is
impossible, even up to a normalizing constant. For example, suppose that we
want to model a chess player's behaviour. We will set up a simple one-parameter
model which describes the chess playing process.  The parameter, $a \in [0, 1]$
indicates the player's eagerness to remove his opponent's pieces from the
board. We can write down an algorithm for the player's behaviour under such a
model.

% TODO: don't break this over the page

\begin{algorithmic}
  \While{the game is not over}
    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
      \State{capture the piece}
    \Else
      \State{make any other move at random}
    \EndIf
  \EndWhile
\end{algorithmic}

Suppose the observed data are the ending configurations of the board, after the
player has concluded a game against an opponent with a known value of $a$. The
model we have designed is very simple, but it is not obvious how to calculate
the likelihood of a particular ending configuration. Indeed, it seems that the
only way is to enumerate every possible path the game could have taken, and
tabulate the ending configurations of each. Clearly, this is infeasible.
Approximate Bayesian computation was designed for situations like these, where
exact likelihoods are not available, perhaps due to the model involving an
algorithm or generative process, but we still want to estimate the parameters.
